Forecasting the 2024 Election: Trump Dominates the South, Harris Leads in the Midwest

### 2024-10-20 paper/paper.qmd
committ changes to paper.qmd

Our model predicts election outcomes by comparing the average polling percentages for Trump and Harris in each state. The results highlight the geographic distribution of support, with some states clearly favoring one candidate over the other. Swing states emerge as critical battlegrounds, where polling percentages are closely contested and could influence the final election result.

@@

ggplot(stateaverage_long, aes(x=reorder(state, -support_pct), y= support_pct, fill = candidate)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black") +
  labs(x = "State", y = "Average Percentage") +
  scale_fill_manual(name = "Candidate", values = c("Trump_pct" = "red", "Harris_pct" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Model {#sec-model}
The goal of our modeling strategy is to predict the winner of the 2024 US Presidential Election in each state based on polling data, while also assessing the likelihood of Donald Trump or Kamala Harris winning. We employ a Bayesian logistic regression model, which allows for probabilistic predictions of binary outcomes, making it well-suited for the task of predicting election results. Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up
The binary outcome variable $y_i$ is defined as 1 if Donald Trump is predicted to win state $i$, and 0 if Kamala Harris is predicted to win. The predictor variables are $\beta_i$, which represents the average percentage of polling support for Donald Trump in state $i$, and $\gamma_i$, the average percentage of support for Kamala Harris in state $i$. The relationship between these variables and the outcome is captured through a logistic link function.

@@

Bayesian logistic regression was chosen for its ability to model binary outcomes like win/loss predictions while incorporating prior knowledge and uncertainty, helping mitigate overfitting through regularization with priors. This model strikes a balance between flexibility and interpretability, given that polling percentages are continuous and the outcome is binary. The model assumes polling data accurately reflects voter intentions and that state outcomes are independent of each other, although it does not account for interactions between states or regional voting trends. Potential biases in polling data could also affect predictions. The model was implemented using the rstanarm package in R, which supports Bayesian inference with default priors for logistic regression. Convergence diagnostics and model checking were conducted to ensure reliable estimates, with further details and alternative models provided in [Appendix -@sec-model-details].

@@

### Summary of Predicted Election Results by State

@@

The logistic regression model used in this analysis successfully captures the relationship between polling percentages and the likelihood of a candidate winning a state. The coefficients for the predictor variables (polling percentages for Trump and Harris) indicate a strong relationship between candidate support and the probability of winning. The model outputs probabilities for each state, which are useful for gauging the certainty of predictions, especially in closely contested states. The model’s performance is reinforced by the robustness of the results in states where one candidate has a clear polling lead, providing accurate predictions. However, states with narrower margins reflect greater uncertainty, which is critical in understanding election dynamics.

## Interpretation of Polling Averages by State

The bar graph displays the average polling percentages for Donald Trump (in red) and Kamala Harris (in blue) across various states. The comparison shows significant variability in support for each candidate. In some states like **Montana** and **Missouri**, Trump’s support is notably higher, with a clear margin over Harris. In contrast, **Colorado** stands out as having nearly equal polling support for both candidates, suggesting a closely contested race in that state.

Swing states such as **Florida**, **North Carolina**, and **Pennsylvania** exhibit relatively close polling percentages, indicating that the election outcomes in these states could be pivotal and difficult to predict. The plot also includes states like **Virginia** and **Texas**, where the support levels are more balanced, though Trump seems to hold a slight lead.

The state labeled as **NA** at the end may indicate missing or incomplete data, emphasizing the need for comprehensive polling coverage in all states for more reliable predictions.

Overall, the side-by-side comparison highlights the states where each candidate holds a significant lead and where the race remains competitive, offering a detailed view of the election's geographic dynamics.

## Weaknesses and next steps

>>>>

Our model utilizes the Bayesian logistic regression model to predict election outcomes by comparing the average polling percentages for Trump and Harris in each state. The results highlight the geographic distribution of support, with some states clearly favoring one candidate over the other. Swing states emerge as critical battlegrounds, where polling percentages are closely contested and could influence the final election result.

@@

ggplot(stateaverage_long, aes(x=reorder(state, -support_pct), y= support_pct, fill = candidate)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black") +
  labs(x = "State", y = "Average Percentage") +
  scale_fill_manual(name = "Candidate", values = c("Trump_pct" = "red", "Harris_pct" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
The comparison shows significant variability in support for each candidate. In some states like **Montana** and **Missouri**, Trump’s support is notably higher, with a clear margin over Harris. In contrast, **Colorado** stands out as having nearly equal polling support for both candidates, suggesting a closely contested race in that state. Swing states such as **Florida**, **North Carolina**, and **Pennsylvania** exhibit relatively close polling percentages, indicating that the election outcomes in these states could be pivotal and difficult to predict. The plot also includes states like **Virginia** and **Texas**, where the support levels are more balanced, though Trump seems to hold a slight lead. The state labeled as **NA** at the end may indicate missing or incomplete data, emphasizing the need for comprehensive polling coverage in all states for more reliable predictions.

# Model {#sec-model}

The goal of our modeling strategy is to predict the winner of the 2024 US Presidential Election in each state based on polling data, while also assessing the likelihood of Donald Trump or Kamala Harris winning. We employ a Bayesian logistic regression model, which allows for probabilistic predictions of binary outcomes, making it well-suited for the task of predicting election results. Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

The binary outcome variable $y_i$ is defined as 1 if Donald Trump is predicted to win state $i$, and 0 if Kamala Harris is predicted to win. The predictor variables are $\beta_i$, which represents the average percentage of polling support for Donald Trump in state $i$, and $\gamma_i$, the average percentage of support for Kamala Harris in state $i$. The relationship between these variables and the outcome is captured through a logistic link function.

@@

Bayesian logistic regression was chosen for its ability to model binary outcomes like win/loss predictions while incorporating prior knowledge and uncertainty, helping mitigate overfitting through regularization with priors. This model strikes a balance between flexibility and interpretability, given that polling percentages are continuous and the outcome is binary. The model assumes polling data accurately reflects voter intentions and that state outcomes are independent of each other, although it does not account for interactions between states or regional voting trends. Potential biases in polling data could also affect predictions. The model was implemented using the rstanarm package in R, which supports Bayesian inference with default priors for logistic regression.

@@

## Summary of Predicted Election Results by State

@@

The logistic regression model used in this analysis successfully captures the relationship between polling percentages and the likelihood of a candidate winning a state. The coefficients for the predictor variables (polling percentages for Trump and Harris) indicate a strong relationship between candidate support and the probability of winning. The model outputs probabilities for each state, which are useful for gauging the certainty of predictions, especially in closely contested states. The model’s performance is reinforced by the robustness of the results in states where one candidate has a clear polling lead, providing accurate predictions. However, states with narrower margins reflect greater uncertainty, which is critical in understanding election dynamics. Convergence diagnostics and model checking were conducted to ensure reliable estimates, with further details and alternative models provided in [Appendix -@sec-model-details].

## Weaknesses and Next Steps
